---
title: "Building Data Visualization Tools (Part 4)"
author: "Pier Lorenzo Paracchini, `r format(Sys.time(), '%d.%m.%Y')`"
output: 
  html_document: 
    fig_height: 5
    fig_width: 7
    keep_md: yes
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE, collapse = TRUE)
```

The content of this blog is based on examples/ notes/ experiments related to the material presented in the "Building Data Visualization Tools" module of the "[Mastering Software Development in R](https://www.coursera.org/specializations/r)" Specialization (Coursera) created by __Johns Hopkins University__ [1].

### Required Packages

* `ggplot2`, a system for 'declaratively' creating graphics, based on "The Grammar of Graphics".
* `gridExtra`, provides a number of user-level functions to work with "grid" graphics.
* `dplyr`, a tool for working with data frame like objects, both in memory and out of memory.
* `viridis`, the viridis color palette.
* `ggmap`, a collection of functions to visualize spatial data and models on top of static maps from various online sources (e.g Google Maps)

```{r packageDependencies}
# If necessary to install a package run
# install.packages("packageName")

# Load packages
library(ggplot2)
library(gridExtra)
library(dplyr)
library(viridis)
library(ggmap)
```

### Data

The `ggplot2` package includes some datasets with geographic information. The `ggplot2::map_data()` function allows to get map data from the `maps` package (use `?map_data` form more information). 

Specifically the `italy` dataset [2] is used for some of the examples below. Please note that this dataset was prepared aroind 1989 so it is out of date especially information pertaing provinces (see `?maps::italy`).

```{r dataExamples}
# Get the italy dataset from ggplot2
# Consider only the following provinces "Bergamo" , "Como", "Lecco", "Milano", "Varese"
# and arrange by group and order (ascending order)
italy_map <- ggplot2::map_data(map = "italy")
italy_map_subset <- italy_map %>%
  filter(region %in% c("Bergamo" , "Como", "Lecco", "Milano", "Varese")) %>%
  arrange(group, order)
```

Each observation in the dataframe defines a geographical point with some extra information:

* `long` & `lat`, longitude and latitude of the geographical point
* `group`, an identifier connected with the specific polygon points are part of
    * a map can be made of different polygons (e.g. one polygon for the main land and one for each islands, one polygon for each state, ...)  
* `order`, the order of the point within the specific `group`
    * how the all of the points being part of the same `group` should be connected in order to create the polygon  
* `region`, the name of the province (Italy) or state (USA) 
    
```{r italyDataExample}
head(italy_map, 3)
```

## How to work with maps 

Having spatial information in the data gives the opportunity to map the data or, in other words, __visualizing the information contained in the data in a geographical context__. R has different possibilities to map data, from normal plots using **longitude**/ **latitude** as `x`/ `y` to more complex spatial data objects (e.g. shapefiles).

### Mapping with `ggplot2` package

The most basic way to create maps with your data is to use `ggplot2`, create a ggplot object and then, add a specific __geom__ mapping **longitute** to `x` aesthetic and **latitude** to `y` aesthetic [4] [5]. This simple approach can be used to:

* create maps of geographical areas (states, country, etc.)  
* map locations as points, lines, etc.

__Create a map showing "Bergamo", Como", "Varese" and "Milano" provinces in Italy using simple points...__

When plotting simple points the `geom_point` function is used. In this case the polygon and order of the points is not important when plotting.

```{r mapItalyExampleAsPoints}
italy_map_subset %>%
  ggplot(aes(x = long, y = lat)) +
  geom_point(aes(color = region))
```

__Create a map showing "Bergamo", Como", "Varese" and "Milano" provinces in Italy using lines...__

The `geom_path` function is used  to create such plots. From the R documentation, `geom_path` _"... connects the observation in the order in which they appear in the data"_. When plotting using `geom_path` is important to consider the polygon and the order within the polygon for each point in the map. 

The points in the dataset are grouped by `region` and ordered by `order`. If information about the region is not provided then the sequential order of the observations will be the order used to connect the points and, for this reason, "unexpected" lines will be drawn when moving from a `region` to the other. On the other hand if information about the region is provided using the `group` or `color` aesthetic, mapping to `region`, the "unexpected" lines are removed (see example below). 


```{r mapItalyExampleAsLines}
plot_1 <- italy_map_subset %>%
  ggplot(aes(x = long, y = lat)) +
  geom_path() +
  ggtitle("No mapping with 'region', unexpected lines")

plot_2 <- italy_map_subset %>%
  ggplot(aes(x = long, y = lat)) +
  geom_path(aes(group = region)) +
  ggtitle("With 'group' mapping")

plot_3 <- italy_map_subset %>%
  ggplot(aes(x = long, y = lat)) +
  geom_path(aes(color = region)) +
  ggtitle("With 'color' mapping")

grid.arrange(plot_1, plot_2, plot_3, ncol = 2, layout_matrix = rbind(c(1,1), c(2,3)))
```

Mapping with `ggplot2` is possible to create more sophisticated maps like choropleth maps [3]. The example below, extracted from [1],  shows how to visualize the percentage of republican votes in 1976 by states.

```{r choroplethExampleUsa}
# Get the USA/ state map from ggplot2
us_map <- ggplot2::map_data("state")

# Use the 'votes.repub' dataset (maps package), containing the percentage of 
# republican votes in the 1900 elections by state. Note
# - the dataset is a matrix so it needs to be converted to a dataframe
# - the row name defines the relevant state 

votes.repub %>%
  tbl_df() %>%
  mutate(state = rownames(votes.repub), state = tolower(state)) %>%
  right_join(us_map, by = c("state" = "region")) %>%
  ggplot(mapping = aes(x = long, y = lat, group = group, fill = `1976`)) +
  geom_polygon(color = "black") + 
  theme_void() +
  scale_fill_viridis(name = "Republican\nVotes (%)")
```

### Maps with `ggmap` package, Google Maps API and others

Another way to create maps is to use the `ggmap`[4] package (see [Google Maps API Terms of Service](http://developers.google.com/maps/terms)). As stated in the package description 

> _"A collection of functions to visualize spatial data and models on top of static maps from various online sources (e.g Google Maps,..). It includes tools common to those tasks, including functions for geolocation and routing."_ R Documentation

The package allows to create/ plot maps using Google Maps and few other service providers, and perform some other interesting tasks like geocoding, routing, distance calculation, etc. The maps are actually __ggplot objects__ making possible to reuse the `ggplot2` functionality like adding layers, modify the theme, ...  

> "The basic idea driving ggmap is to take a downloaded map image, plot it as a context layer using ggplot2, and then plot additional content layers of data, statistics, or models on top of the map. In ggmap this process is broken into two pieces – (1) downloading the images and formatting them for plotting, done with get_map, and (2) making the plot, done with ggmap. qmap marries these two functions for quick map plotting (c.f. ggplot2’s ggplot), and qmplot attempts to wrap up the entire plotting process into one simple command (c.f. ggplot2’s qplot)." [4]

__How to create and plot a map...__

The `ggmap::get_map`function is used to get a base map (a `ggmap` object, a raster object)  from different service providers like Google Maps, OpenStreetMap, Stamen Maps or Naver Maps (default setting is Google Maps). Once the base map is available, then it can been plotted using the `ggmap::ggmap` function. Alternatively the `ggmap::qmap` function (quick map plot) can be used.

```{r vareseMap, collapse=TRUE}
# When querying for a base map the location must be provided
# name, address (geocoding)
# longitude/ latitude pair
base_map <- get_map(location = "Varese")
ggmap(base_map) + ggtitle("Varese")
```

```{r vareseMap_a, collapse=TRUE}
# qmap is a wrapper for 
# `ggmap::get_map` and `ggmap::ggmap` functions.
qmap("Varese") + ggtitle("Varese - qmap")
```


__How to change the zoom in the map...__

The `zoom` argument (default value is `auto`) in `ggmap::get_map` function can be  used to control the zoom of the returned base map (see `?get_map` for more information). Please note that the possible values/ range for the `zoom` argument changes with the different sources. 

```{r vareseMapWithZooms, collapse = TRUE}
# An example using Google Maps as a source
# Zoom is an integer between 3 - 21 where
# zoom = 3 (continent)
# zoom = 10 (city)
# zoom = 21 (building)

base_map_10 <- get_map(location = "Varese", zoom = 10)
base_map_18 <- get_map(location = "Varese", zoom = 16)

grid.arrange(ggmap(base_map_10) + ggtitle("Varese, zoom 10"), 
             ggmap(base_map_18) + ggtitle("Varese, zoom 18"), 
             nrow = 1)
```

__How to change the type of map...__

The `maptype` argument in `ggmap::get_map` function can be used to change the type of map aka __map theme__. Based on the R documentation (see `?get_map` for more information)

> '[maptype]... options available are "terrain", "terrain-background", "satellite", "roadmap", and "hybrid" (google maps), "terrain", "watercolor", and "toner" (stamen maps)...'.

```{r vareseMapWithMapTypes, collapse = TRUE}
# An example using Google Maps as a source
# and different map types

base_map_ter <- get_map(location = "Varese", maptype = "terrain")
base_map_sat <- get_map(location = "Varese", maptype = "satellite")
base_map_roa <- get_map(location = "Varese", maptype = "roadmap")

grid.arrange(ggmap(base_map_ter) + ggtitle("Terrain"), 
             ggmap(base_map_sat) + ggtitle("Satellite"), 
             ggmap(base_map_roa) + ggtitle("Road"),
             nrow = 1)
```

__How to change the source for maps...__

While the default source for maps with `ggmap::get_map` is Google Maps, it is possible to change the map service using the `source` argument. The supported map services/ sources are Google Maps, OpenStreeMaps, Stamen Maps and CloudMade Maps (see `?get_map` for more information).

```{r mapWithSources, collapse = TRUE}
# An example using different map services as a source

base_map_google <- get_map(location = "Varese", source = "google", maptype = "terrain")
base_map_stamen <- get_map(location = "Varese", source = "stamen", maptype = "terrain")

grid.arrange(ggmap(base_map_google) + ggtitle("Google Maps"), 
             ggmap(base_map_stamen) + ggtitle("Stamen Maps"), 
             nrow = 1)
```

__How to geocode a location...__

The `ggmap::geocode` function can be used to find latitude and longitude of a location based on its name (see `?geocode` for more information). Note that Google Maps API limits the possible number of queries per day, `geocodeQueryCheck` can be used to determine how many queries are left.

```{r geocoding, collapse=TRUE}
# Geocode a city
geocode("Sesto Calende")
# Geocode a set of cities
geocode(c("Varese", "Milano"))

# Geocode a location
geocode(c("Milano", "Duomo di Milano"))
geocode(c("Roma", "Colosseo"))
```

__How to find a route between two locations...__

The `ggmap::route` function can be used to find a route from Google using different possible modes, e.g. walking, driving, ... (see `?ggmap::route` for more information).

>'The route function provides the map distances for the sequence of "legs" which constitute a route between
two locations. Each leg has a beginning and ending longitude/latitude coordinate along with a distance and duration in the same units as reported by mapdist. The collection of legs in sequence constitutes a single route (path) most easily plotted with geom_leg, a new exported ggplot2 geom...' [4]

```{r findRouteExample, collapse = TRUE}
route_df <- route(from = "Somma Lombardo", to = "Sesto Calende", mode = "driving")
head(route_df)
```

```{r routingExample, collapse = TRUE}
route_df <- route(from = "Via Gerolamo Fontana 32, Somma Lombardo", 
                  to = "Town Hall, Somma Lombardo", mode = "walking")

qmap("Somma Lombardo", zoom = 16) +
  geom_leg(
    aes(x = startLon, xend = endLon, y = startLat, yend = endLat),  colour = "red",
    size = 1.5, alpha = .5,
    data = route_df) + 
  geom_point(aes(x = startLon, y = startLat), data = route_df) +
  geom_point(aes(x = endLon, y = endLat), data = route_df)
```


__How to find the distance between two locations...__

The `ggmap::mapdist` function can be used to compute the distance between two location using different possible modes, e.g. walking, driving, ... (see `?ggmap::mapdist` for more information).

```{r findDistanceExample, collapse = TRUE}
# Driving
mapdist(from = "Somma Lombardo", to = "Sesto Calende", mode = "driving")
# Walking
mapdist(from = "Somma Lombardo", to = "Sesto Calende", mode = "walking")
```

## More on mapping

- Using the `choroplethr` and `choroplethrMaps` packages, see "Mapping US counties and states" section in [1]   
- Working with spatial objects and shapefiles, see "More advanced mapping – Spatial objects" section in [1]
- Using htmlWidgets for mapping in R using __leaflet__ [5]

# References

[1] "Mapping" chapter in "[Mastering Software Development in R](https://bookdown.org/rdpeng/RProgDA/mapping.html/)" by Roger D. Peng, Sean Cross and Brooke Anderson, 2017  
[2] Italy Map, "UNESCO (1987) through UNEP/GRID-Geneva"  
[3] Choropleth map [Wikipedia](https://en.wikipedia.org/wiki/Choropleth_map)  
[4] D. Kahle and H. Wickham. [ggmap: Spatial Visualization with ggplot2](http://journal.r-project.org/archive/2013-1/kahle-wickham.pdf). The R Journal, 5(1), 144-161.  
[5] Using [Leaflet](https://rstudio.github.io/leaflet/) for R

## Previous "Building Data Visualization Tools" blogs

[6] "[Basic plotting with R and ggplot2](https://pparacch.github.io/2017/07/06/plotting_in_R_ggplot2_part_1.html)", Part 1  
[7] "['ggplot2', essential concepts](https://pparacch.github.io/2017/07/14/plotting_in_R_ggplot2_part_2.html)", Part 2  
[8] "[Guidelines for good plots](https://pparacch.github.io/2017/07/18/plotting_in_R_ggplot2_part_3.html)", Part 3

